---
title: "Introduction to the penppml package"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to the penppml package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(penppml)
```

# Introduction
`penppml` is an R package that enables you to fit penalized Poisson Pseudo Maximum Likelihood models with high-dimensional fixed effects. Supported penalties in the current version are ridge and lasso. The original application that motivated the development of `penppml` was the estimation of gravity models of trade with a large number of PTA provision dummies (Breinlich, Corradi, Rocha, Ruta, Santos Silva and Zylkin, 2021).

# Main functions
The workhorse of our package is the `mlfitppml` function.

# Applications 
## The effects of specific trade agreement provisions on international trade

# References
