---
title: "Penalized PPML Regression with penppml: an introduction"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Penalized PPML Regression with penppml: an introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(penppml)
```

# Introduction
`penppml` is an R package that enables you to fit penalized Poisson Pseudo Maximum Likelihood models with high-dimensional fixed effects. Supported penalties in the current version are ridge and lasso. The original application that motivated the development of `penppml` was the estimation of gravity models of trade with a large number of PTA provision dummies (Breinlich, Corradi, Rocha, Ruta, Santos Silva and Zylkin, 2021).

# Main functions
The workhorse of our package is the `mlfitppml` function.

# Applications: the effects of specific trade agreement provisions on international trade

The `penppml` package features the `trade` data set, which integrates panel data on bilateral trade flows with information about specific provisions in trade agreements for 220 exporters and 270 importers. The provisions included in the package are a subset of 50 out of 305 "essential" provisions featured in the full data set. More information about the data set and the variables is included in the corresponding help file, accessible via `?trade`.

```{r inspect data, echo = FALSE}
knitr::kable(head(trade[, c(1:4, 10:13)], 10), format = "pipe")
```

We are going to use the `trade` data set to demonstrate the capabilities of this package. First, we will run an unpenalized PPML regression using `hdfeppml`:

```{r try hdfeppml}
reg <- hdfeppml2(data = trade[, -(5:9)],
          dep = "export",
          fixed = c("exp", "imp", "time"),
          interactions = list(c("exp", "time"), 
                              c("imp", "time"),
                              c("exp", "imp")))
```

```{r display results, echo = FALSE}
results <- data.frame(prov = rownames(reg$coefficients), b = reg$coefficients, se = 0)
results$se[!is.na(reg$coefficients)] <- reg$se
knitr::kable(list(results[1:13, ], results[14:26, ], 
                  results[27:39, ], results[40:50, ]), 
             format = "pipe",
             col.names = c("Provision", "Coefficient", "SE"), 
             caption = "Table 1: Unpenalized PPML results",
             row.names = FALSE,
             digits = 4)
```


# References
